Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-16 MScDiplom_Lemenkova.tex
Comment 35 of Valentijn: Reported Classification accuracy and printed Kappa confusion matrix. Added two paragraphs: 'Data inregration' and 'Accuracy assessment'

\documentclass[11pt]{article}
\special{papersize=140mm,210mm}
\usepackage{authblk}
\usepackage[english]{babel}
%\usepackage{authorindex}
\usepackage{color}
\usepackage[stable]{footmisc }
\usepackage{setspace}
\usepackage[printonlyused,withpage]{acronym}
\usepackage[left=30mm,right=20mm,top=30mm,bottom=35mm]{geometry}
\usepackage{pslatex}
%\usepackage{times}
%\usepackage{psnfss}
%\usepackage{glossary}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
%\usepackage{makeidx}\makeindex
%\usepackage[makeindex,protected]{splitidx}
%\usepackage{index}
\usepackage{imakeidx}
\makeindex[name=concepts,title=Index of Concepts,intoc]
\makeindex[name=people,title=Index of People,intoc,columns=3,columnseprule]
\usepackage[fit]{truncate}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[RO,LE]{\truncate{.95\headwidth}{\leftmark}}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage{hyperref}
\hypersetup{pdftitle={Seagrass mapping and monitoring along the coasts of Crete, Greece}, pdfauthor={Lemenkova Polina}, pdfsubject={MScThesis}, pdfcreator={Lemenkova Polina}, pdfproducer={Lemenkova Polina}, colorlinks=true,linkcolor=Navy, citecolor=Crimson, filecolor=magenta, urlcolor=blue,pdfkeywords={P.oceanica, seagrass, spectral reflectance, remote sensing, WASI, RTM, Google Earth, mapping, Crete}}

@@

\subsection{Integrated mapping of seagrass}

\index[concepts]{Data!integration}
The integrated approach used in this research work has high potential as a means to monitor changes
in seagrass landscape occurring in shallow waters over Crete area.

@@

It encompasses the integration of
high resolution aerial color Google Earth photography, spaceborne satellite imagery, assessment of
spectral signatures using \ac{WASI} software\index[concepts]{Seawater optical properties!RTM!WASI}, image processing by means of Erdas Imagine (Fig.\ref{fig:39}, Fig.\ref{fig:42}) and Arc \ac{GIS} based mapping. The use of \ac{GIS} for data incorporation (Fig.\ref{fig:4.17}), storage, analyses, visualizing and mapping
enables to analyze environmental changes within seagrass landscapes based on data from various
sources: aerial and satellite images, geographically referenced maps of Crete island and results of
images classification showing areas of seagrass distribution.
The final mapping has been supported in ArcGIS through the data exporting, conversion and
integration of various data in one \ac{GIS}-project. Data collected during the fieldwork, imagery
of the seagrass distribution are added into a \ac{GIS} dataset for the assessment and spatial analysis.
\pagebreak

@@

Besides spatial structure of the seagrass meadows, there are strong and complex patterns of depth zoning, specific to individual
seagrass species, but such detailed classification goes beyond the scope of the current study.
Unique feature of seagrass meadows, most characteristic for shallow areas, is dynamics of their
landscapes. Homogeneous, continuous seagrass meadows can reach in size up to 40 m2 (Fig.\ref{fig:38} a),
however they are often interrupted by gaps and channels of open spaces generated by the complex,
disturbing sedimentation processes, turbulence and turbidity regime of waters and landscape dynamics,
which form gaps and channels within meadows, making them “patchy-looking”, i.e. diversified by
separate mattes (Fig.\ref{fig:38} b). The traditional definition for gap in vegetation cover is “disturbance generated
openings in either floral or faunal cover” \cite{Connell78}\label{Connell78}.
Formation and increase of these
gaps within the mosaic of seagrass meadows is caused by different reasons. The most probable
drivers for the process of gaps formation within seagrass meadows are removal of interior vegetation,
differential growth of seagrass meadows and increased sedimentation. Thus, storms lead to severe
deposition of sediments, burying parts of seagrass meadow, the same effect has movements and
deposition of sediments during and after floods\cite{Bell99}\label{Bell99}.
Increased nutrient sedimentation, especially phosphorus, were explored \cite{Jensen01}\label{Jensen01} as a
potential mechanism for increasing patch dynamics and morphological plasticity within seagrass
meadows. Finally, increasing the degree of fragmentation of the landscapes of 	extit{P.oceanica} meadows
can be caused by the invasion of alien species, such as \textit{Cymodocea nodosa, Caulerpa prolifera,
Caulerpa taxifolia}. Invaders are in general strong colonizers comparing to native \textit{P.oceanica}: they
occupy much greater habitat space within the regressed meadows of stressed native seagrass
\cite{Montefalcone10}\label{Montefalcone10}. However, on the northern coasts of Crete the only dominating seagrass
species is \textit{P.oceanica}.
Morphological differences in scale of seagrass landscape formations, discussed above, cause need for
the different-scale mapping. Therefore, the investigation of the seagrass meadows at different levels is
performed using underwater videometric measurements, aerial and satellite imagery.

\section{Conclusion}\label{sec:6}
The goal of this MSc research was to explore the perspectives, advantages and limitations of the narrow-band and broadband sensors for the environmental mapping and monitoring of \textit{P.oceanica} seagrass along the coasts of Crete Island. The research outcome demonstrated that the application of the remote sensing data from the broadband sensors is highly advantageous for the seagrass mapping, the spectral discrimination of \textit{P.oceanica} from other seafloor cover types is possible at diverse and changing environmental conditions, and that \textit{P.oceanica} is spectrally distinct from other seagrass species as well (\textit{Thalassia testudinum}), Fig.\ref{fig:2.6}\label{page-54}.\vspace{1ex}\\
The \ac{RTM} software is a powerful means for analyzing spectral signatures of various seafloor types and enabling simulations of data received from broadband and narrowband remote sensors.\\ The example of application of \ac{WASI} \ac{RTM}, given in this work, is an achievement of the research insight towards the spectral properties of \textit{P.oceanica} and other bottom cover types, enabling to discriminate them from each other with changing environmental conditions. The research shows that spectral signatures of \textit{P.oceanica} are distinct at various depths.\vspace{1ex}\\
The methodology of the spectral discrimination of seafloor cover types is designed in the frame of this research and is based on the application of the remote sensing \ac{RTM} techniques, data from broadband sensors, hyperspectral radiometers for measurements of optical properties of the seawater, categorical and continuous statistical analysis for the data processing and \ac{GIS} raster based software for images visualization, classification and analysis. Technically, we used different software, adjusted for diverse research purposes, to manage, integrate and process data from various origin and resources, and finally to receive  accurate research results.\vspace{1ex}\\
The marine coastal ecosystems are complex, constantly changing and developing. Using flexibility of \ac{GIS} combined with remote sensing methods and application of data from broadband sensors is therefore advantageous for the monitoring of coastal areas. Besides Mediterranean area, the methodology of the seagrass environmental studies can be applied towards other shelf areas with dominating seagrass landscapes.\vspace{1ex}\\
More than 50 \% of the world population lives within one km of the coast, which results in continued anthropogenic pressure on the coastal regions. Therefore, management of coastal resources and shelf zone protection become increasingly important nowadays, and require large-scale monitoring and mapping of the shelf areas as a vital instrument for the environmental assessment.\vspace{1ex}\\
This research is a contribution to the development of the methodology of seagrass mapping with the aim of the environmental monitoring, and a case study of \textit{P.oceanica}, an endemic Mediterranean seagrass species, dominating in underwater landscapes and ecosystems along the coasts of Crete Island.

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.23]{OpenEV-2.jpg}
		\caption{The point querying shows the selected points and their coordinates within the area of seagrass meadow (green). \href{http://openev.sourceforge.net/}{OpenEV}}\label{fig:A.39}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.23]{OpenEV-3.jpg}
	\caption{\textit{Logarithmic Enhancement to Raster}, applied to the aerial Google Earth image: the seagrass meadow seen as a bright spot of purple color. \href{http://openev.sourceforge.net/}{OpenEV}}\label{fig:A.40}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.25]{OpenEV-1.jpg}
		\caption{Raster properties dialog: visualization and spatial info about the image (projection UTM, zone 35, datum WGS-84, etc.)  \href{http://openev.sourceforge.net/}{OpenEV})}\label{fig:A.41}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.25]{Ilwis-2.jpg}
		\caption{Color composite image composed of 3 images of Cretan shelf, Google Earth. \href{http://www.ilwis.org/}{ILWIS}}\label{fig:A.42}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.25]{Fig-39.jpg}
		\caption{Results of the unsupervised classification of the seafloor cover types and land structure, Agia Pelagia; raster layer read into the \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS} project. Page: $<$\pageref{page-48}$>$}
		\label{fig:42}
	\end{center}
\end{figure}
\begin{figure}
	\begin{center}
		\includegraphics[scale=0.25]{Fig-36.jpg}
		\caption{Results of the image classification in \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine}: \\ seagrass distribution in Bali area, Crete. Page: $<$\pageref{page-48}$>$}
		\label{fig:39}
	\end{center}
\end{figure}
\pagebreak

>>>>

\documentclass[11pt]{article}
\special{papersize=140mm,210mm}
\usepackage{authblk}
\usepackage[english]{babel}
%\usepackage{authorindex}
\usepackage{color}
\usepackage[stable]{footmisc}
\usepackage{setspace}
\usepackage[printonlyused,withpage]{acronym}
\usepackage[left=30mm,right=20mm,top=30mm,bottom=35mm]{geometry}
\usepackage{pslatex}
%\usepackage{times}
%\usepackage{psnfss}
%\usepackage{glossary}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage{rotating}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
%\usepackage{makeidx}\makeindex
%\usepackage[makeindex,protected]{splitidx}
%\usepackage{index}
\usepackage{imakeidx}
\makeindex[name=concepts,title=Index of Concepts,intoc]
\makeindex[name=people,title=Index of People,intoc,columns=3,columnseprule]
\usepackage[fit]{truncate}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[RO,LE]{\truncate{.95\headwidth}{\leftmark}}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage{hyperref}
\hypersetup{pdftitle={Seagrass mapping and monitoring along the coasts of Crete, Greece}, pdfauthor={Lemenkova Polina}, pdfsubject={MScThesis}, pdfcreator={Lemenkova Polina}, pdfproducer={Lemenkova Polina}, colorlinks=true,linkcolor=Navy, citecolor=Crimson, filecolor=magenta, urlcolor=blue,pdfkeywords={P.oceanica, seagrass, spectral reflectance, remote sensing, WASI, RTM, Google Earth, mapping, Crete}}

@@

\subsection{GIS mapping of seagrass}

\index[concepts]{Data!integration}
\paragraph{Data integration}
The integrated approach used in this research work has high potential as a means to monitor changes
in seagrass landscape occurring in shallow waters over Crete area.

@@

It encompasses the integration of
high resolution aerial color Google Earth photography, spaceborne satellite imagery, assessment of
spectral signatures using \ac{WASI} software\index[concepts]{Seawater optical properties!RTM!WASI}, image processing by means of Erdas Imagine (Fig.\ref{fig:39}, Fig.\ref{fig:42}) and Arc \ac{GIS} based mapping. The use of \ac{GIS} for data incorporation (Fig.\ref{fig:4.17}), storage, analyses, visualizing and mapping
enables to analyze environmental changes within seagrass landscapes based on data from various
sources: aerial and satellite images, geographically referenced maps of Crete island and results of
images classification showing areas of seagrass distribution.
The final mapping has been supported in ArcGIS through the data exporting, conversion and
integration of various data in one \ac{GIS}-project. Data collected during the fieldwork, imagery
of the seagrass distribution are added into a \ac{GIS} dataset for the assessment and spatial analysis.
\paragraph{Accuracy assessment}\label{page-50acc}
We prepared the error confusion matrices (Fig.\ref{tab:kappa-1} and Fig.\ref{tab:kappa-2}) using $\kappa$ (Kappa-) statistics to assess and evaluate accuracy of the classification. The accuracy assessment has been done using Erdas Imaging function Classifier/Accuracy Assessment. In the Accuracy Assessment viewer we have chosen utility Edit in order to generate random points throughout the classified image, and then chosen the Create or add random points dialog. After the points were generated, we entered the class values for the reference points from the supervised classification. In order to perform a proper accuracy assessment we needed about 300 points so that we defined the number of points for the selection. From the option parameters for points distribution we have chosen "stratified random" as better representing the variables. Then we evaluated the location of the points and determined their class value, which was done with ground truth points from Google Earth aerial photos for various data and seafloor imagery and with reference to the signature file, to find the numeric value that has been assigned to each class previously. Finally, to print the errors matrix, we selected Accuracy Assessment viewer again and chosen there Report/ Options with turned on Error Matrix, Kappa Statistics and Accuracy Total. The error matrix is just comparing reference points of various seafloor cover types to the classified points (i.e. seagrass, carbonate sand, various land cover types on the coast). The Kappa coefficient takes into account chance agreement and thus, shows the reduction in errors generated by a classification process compared with the errors which could be received by a completely random classification – in other words, it evaluates quality of the classification. The results of the classification can be seen on the Fig. (see appendices). The overall map accuracy is 68\%, which means that 68\% of the pixels are classified to the correctly chosen seafloor cover type.
\pagebreak

@@

Besides spatial structure of the seagrass meadows, there are strong and complex patterns of depth zoning, specific to individual
seagrass species, but such detailed classification goes beyond the scope of the current study.
Unique feature of seagrass meadows, most characteristic for shallow areas, is dynamics of their
landscapes. Homogeneous, continuous seagrass meadows can reach in size up to 40 m2 (Fig.\ref{fig:38} a),
however they are often interrupted by gaps and channels of open spaces generated by the complex,
disturbing sedimentation processes, turbulence and turbidity regime of waters and landscape dynamics,
which form gaps and channels within meadows, making them “patchy-looking”, i.e. diversified by
separate mattes (Fig.\ref{fig:38} b). The traditional definition for gap in vegetation cover is “disturbance generated
openings in either floral or faunal cover” \cite{Connell78}\label{Connell78}.
Formation and increase of these
gaps within the mosaic of seagrass meadows is caused by different reasons. The most probable
drivers for the process of gaps formation within seagrass meadows are removal of interior vegetation,
differential growth of seagrass meadows and increased sedimentation. Thus, storms lead to severe
deposition of sediments, burying parts of seagrass meadow, the same effect has movements and
deposition of sediments during and after floods\cite{Bell99}\label{Bell99}.
Increased nutrient sedimentation, especially phosphorus, were explored \cite{Jensen01}\label{Jensen01} as a
potential mechanism for increasing patch dynamics and morphological plasticity within seagrass
meadows. Finally, increasing the degree of fragmentation of the landscapes of 	extit{P.oceanica} meadows
can be caused by the invasion of alien species, such as \textit{Cymodocea nodosa, Caulerpa prolifera,
Caulerpa taxifolia}. Invaders are in general strong colonizers comparing to native \textit{P.oceanica}: they
occupy much greater habitat space within the regressed meadows of stressed native seagrass
\cite{Montefalcone10}\label{Montefalcone10}. However, on the northern coasts of Crete the only dominating seagrass
species is \textit{P.oceanica}.
Morphological differences in scale of seagrass landscape formations, discussed above, cause need for
the different-scale mapping. Therefore, the investigation of the seagrass meadows at different levels is
performed using underwater videometric measurements, aerial and satellite imagery.
\pagebreak

\section{Conclusion}\label{sec:6}
The goal of this MSc research was to explore the perspectives, advantages and limitations of the narrow-band and broadband sensors for the environmental mapping and monitoring of \textit{P.oceanica} seagrass along the coasts of Crete Island. The research outcome demonstrated that the application of the remote sensing data from the broadband sensors is highly advantageous for the seagrass mapping, the spectral discrimination of \textit{P.oceanica} from other seafloor cover types is possible at diverse and changing environmental conditions, and that \textit{P.oceanica} is spectrally distinct from other seagrass species as well (\textit{Thalassia testudinum}), Fig.\ref{fig:2.6}\label{page-54}.\vspace{1ex}\\
The \ac{RTM} software is a powerful means for analyzing spectral signatures of various seafloor types and enabling simulations of data received from broadband and narrowband remote sensors.\\ The example of application of \ac{WASI} \ac{RTM}, given in this work, is an achievement of the research insight towards the spectral properties of \textit{P.oceanica} and other bottom cover types, enabling to discriminate them from each other with changing environmental conditions. The research shows that spectral signatures of \textit{P.oceanica} are distinct at various depths.\vspace{1ex}\\
The methodology of the spectral discrimination of seafloor cover types is designed in the frame of this research and is based on the application of the remote sensing \ac{RTM} techniques, data from broadband sensors, hyperspectral radiometers for measurements of optical properties of the seawater, categorical and continuous statistical analysis for the data processing and \ac{GIS} raster based software for images visualization, classification and analysis. Technically, we used different software, adjusted for diverse research purposes, to manage, integrate and process data from various origin and resources, and finally to receive  accurate research results.\vspace{1ex}\\
The marine coastal ecosystems are complex, constantly changing and developing. Using flexibility of \ac{GIS} combined with remote sensing methods and application of data from broadband sensors is therefore advantageous for the monitoring of coastal areas. Besides Mediterranean area, the methodology of the seagrass environmental studies can be applied towards other shelf areas with dominating seagrass landscapes.\vspace{1ex}\\
More than 50 \% of the world population lives within one km of the coast, which results in continued anthropogenic pressure on the coastal regions. Therefore, management of coastal resources and shelf zone protection become increasingly important nowadays, and require large-scale monitoring and mapping of the shelf areas as a vital instrument for the environmental assessment.\vspace{1ex}\\
This research is a contribution to the development of the methodology of seagrass mapping with the aim of the environmental monitoring, and a case study of \textit{P.oceanica}, an endemic Mediterranean seagrass species, dominating in underwater landscapes and ecosystems along the coasts of Crete Island.
\pagebreak

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.23]{OpenEV-2.jpg}
		\caption{The point querying shows the selected points and their coordinates within the area of seagrass meadow (green). \href{http://openev.sourceforge.net/}{OpenEV}}\label{fig:A.39}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.23]{OpenEV-3.jpg}
	\caption{\textit{Logarithmic Enhancement to Raster}, applied to the aerial Google Earth image: the seagrass meadow seen as a bright spot of purple color. \href{http://openev.sourceforge.net/}{OpenEV}}\label{fig:A.40}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.25]{OpenEV-1.jpg}
		\caption{Raster properties dialog: visualization and spatial info about the image (projection UTM, zone 35, datum WGS-84, etc.)  \href{http://openev.sourceforge.net/}{OpenEV})}\label{fig:A.41}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.25]{Ilwis-2.jpg}
		\caption{Color composite image composed of 3 images of Cretan shelf, Google Earth. \href{http://www.ilwis.org/}{ILWIS}}\label{fig:A.42}	
	\end{center}	
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.25]{Fig-39.jpg}
		\caption{Results of the image unsupervised classification of the seafloor cover types and land structure, Agia Pelagia; raster layer read into the \href{http://www.esri.com/software/arcgis/index.html}{ArcGIS} project. Page: $<$\pageref{page-48}$>$}
		\label{fig:42}
	\end{center}
\end{figure}
\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.25]{Fig-36.jpg}
		\caption{Results of the image supervised classification in \href{http://www.erdas.com/products/ERDASIMAGINE/ERDASIMAGINE/Details.aspx}{Erdas Imagine}: \\ seagrass distribution in Bali area, Crete. Pages: $<$\pageref{page-48}$>$, $<$\pageref{page-liii}$>$}
		\label{fig:39}
	\end{center}
\end{figure}
\pagebreak

\subsection{Accuracy assessment}

\begin{table}[H]\footnotesize\label{page-liii}
	\caption{Confusion Matrix-1, between the classified Google Earth aerial image and fieldwork data, for Fig.\ref{fig:42}. Page: $<$\pageref{page-50acc}$>$}\label{tab:kappa-1}
	\centering
	\rowcolors{1}{Honeydew}{White}
	 \begin{tabular}{| p{1.7cm} | p{0.4cm}| p{0.4cm}|p{0.4cm}|p{0.4cm}|p{0.4cm}|p{0.4cm}|p{0.4cm}|p{0.4cm}|p{0.4cm}|p{0.4cm}|p{0.4cm}|p{0.4cm}|p{0.4cm}| p{0.4cm}| p{0.4cm}|| p{0.4cm}| p{0.4cm}|}
		     \textbf{Correctly classified} & \begin{sideways}\textbf{Rocky bottom}\end{sideways} &  \begin{sideways}\textbf{Shelf, $<3$m}\end{sideways} &  \begin{sideways}\textbf{Healthy seagrass}\end{sideways} &  \begin{sideways}\textbf{Seagrass \textit{P.oceanica}}\end{sideways} &  \begin{sideways}\textbf{Carbonate sand}\end{sideways} &  \begin{sideways}\textbf{Shelf waters, 0-3m}\end{sideways} &  \begin{sideways}\textbf{Shelf waters, 3-7m}\end{sideways} & \begin{sideways}\textbf{Deep waters, $>7$m}\end{sideways} &  \begin{sideways}\textbf{Field: corn, greens}\end{sideways} & \begin{sideways}\textbf{Roads: asphalt+ground}\end{sideways} & \begin{sideways}\textbf{Seagrass, other$>3$m}\end{sideways} & \begin{sideways}\textbf{Ground}\end{sideways}& \begin{sideways}\textbf{Buildings (roofs)}\end{sideways} &\begin{sideways}\textbf{Trees}\end{sideways} & \begin{sideways}\textbf{Bushes}\end{sideways} & \begin{sideways}\textbf{Total}\end{sideways}& \begin{sideways}\textbf{$\kappa$ producer accuracy}\end{sideways} \\ \hline \hline
		\textbf{Rocky bottom}& \cellcolor{GreenYellow}\textbf{14} & 2 & 0 & 0 & 3 & 0 & 0 & 0 & 0 & 1 & 0 & 2 & 0 & 1 & 1 & \cellcolor{LightGoldenrodYellow}24 & 0.58 \\ \hline
		\textbf{Shelf, $<3$m} & 0 & \cellcolor{GreenYellow}\textbf{12} & 0 & 3 & 1 & 2 & 0& 0 & 1 & 0 & 0 & 0 & 1 & 0 & 0 & \cellcolor{LightGoldenrodYellow}20 & 0.60 \\ \hline
		\textbf{Healthy seagrass} & 0 & 1 & \cellcolor{GreenYellow}\textbf{10} & 3 & 2 & 0 & 0 & 0 & 1 & 0 & 2 & 0 & 0 & 0 & 0 & \cellcolor{LightGoldenrodYellow}19 & 0.52 \\ \hline
             	\textbf{Seagrass \textit{P.oceanica}} & 0 & 1 & 1 & \cellcolor{GreenYellow}\textbf{13} & 0 & 1 & 0 & 0 & 1 & 0 & 2 & 0 & 0 & 1 & 1 & \cellcolor{LightGoldenrodYellow}21 & 0.62 \\ \hline
		\textbf{Carbonate sand} & 2 & 0 & 0 & 0 & \cellcolor{GreenYellow}\textbf{16} & 0 & 0 & 0 & 0 & 0 & 0 & 2 & 0 & 0 & 0 & \cellcolor{LightGoldenrodYellow}20 & 0.80 \\ \hline
		\textbf{Shelf waters, 0-3m} & 0 & 3 & 0 & 0 & 0 & \cellcolor{GreenYellow}\textbf{9} & 2 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & \cellcolor{LightGoldenrodYellow}14 & 0.64 \\ \hline
		\textbf{Shelf waters, 3-7m} & 1 & 2 & 0 & 0 & 0 & 3 & \cellcolor{GreenYellow}\textbf{8} & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & \cellcolor{LightGoldenrodYellow}15 & 0.53 \\ \hline
		\textbf{Deep waters, $>7$m} & 1 & 1 & 0 & 0 & 1 & 2 & 3 & \cellcolor{GreenYellow}\textbf{16} & 0 & 0 & 1 & 0 & 0 & 0 & 0 & \cellcolor{LightGoldenrodYellow}25 & 0.64 \\ \hline
		\textbf{Fields: corn, greens} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & \cellcolor{GreenYellow}\textbf{14} & 0 & 0 & 0 & 0 & 3 & 4 & \cellcolor{LightGoldenrodYellow}21 & 0.66 \\ \hline
		\textbf{Roads: asphalt+ground} & 1 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & \cellcolor{GreenYellow}\textbf{5} & 2 & 0 & 0 & 0 & 0 & \cellcolor{LightGoldenrodYellow}9 & 0.55 \\ \hline
		\textbf{Seagrass, other$>3$m} & 0 & 0 & 0 & 3 & 0 & 0 & 0 & 0 & 0 & 0 & \cellcolor{GreenYellow}\textbf{24} & 2 & 0 & 1 & 1 & \cellcolor{LightGoldenrodYellow}31 & 0.77 \\ \hline		
		\textbf{Ground} & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & \cellcolor{GreenYellow}\textbf{11} & 3 & 1 & 1 & \cellcolor{LightGoldenrodYellow}17 & 0.65 \\ \hline
		\textbf{Buildings (roofs)} & 0 & 0 & 0 & 0 & 2 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & \cellcolor{GreenYellow}\textbf{9} & 2 & 3 &\cellcolor{LightGoldenrodYellow} 16 & 0.56 \\ \hline
		\textbf{Trees} & 0 & 0 & 2 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & \cellcolor{GreenYellow}\textbf{7} & 3 & \cellcolor{LightGoldenrodYellow}12 & 0.58 \\ \hline
		\textbf{Bushes} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 2 & \cellcolor{GreenYellow}\textbf{4}& \cellcolor{LightGoldenrodYellow}6 & 0.66 \\ \hline	\hline	
		\rowcolor{LightGoldenrodYellow}\textbf{Total} & 20 & 22 & 13 & 22 & 26 & 17 & 13 & 17 & 17 & 6 & 31 & 17 & 13 & 18 & 18 & \cellcolor{LightGoldenrodYellow}\textbf{270} & \\ \hline
		\textbf{$\kappa$ user`s accuracy} & 0.70 & 0.55 & 0.77 & 0.59 & 0.62 & 0.53 & 0.62 & 0.94 & 0.82 & 0.83 & 0.77 & 0.65 & 0.69 & 0.39 & 0.22 & - & \cellcolor{Chartreuse}\textbf{0.64} \\ \hline
\normalsize
Overall Kappa ($\kappa$) accuracy is calculated using the formula:  $\sum{A/N}$, where A is number of correctly mapped points (172) and N is the total number of points (270). Thus, according to the results we received overall accuracy= 172/270= 0.6370, which is 64\%.  Overall $\kappa$ accuracy for unsupervised classification =\textbf{64\%}. \\ 
 User’s accuracy (Reliability of classes) varies between 0.22 and 0.94 depending on class, which proves that supervised classification (see next table: Tab.\ref{tab:kappa-2} ) has better results for seagrass mapping than the unsupervsied classification. 
Producer accuracy lies in interval between 0.52-0.77 according to class as well. 
